Tissue Enrichment Analysis (TEA)

### 2016-04-14 tests/hgf_benchmarking/doc/tea_main_text.tex
Changed figure format from pdf to png
Version change from 13.9 -> 13.10
Figures are now saved as .pdf

Some files reformatted to be written in PEP8.

Writing.

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes. As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}. These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}). A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.png}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.}
	  \label{fig:qvals}
\end{figure}

@@

%gaba
\begin{figure}
  \includegraphics[width=0.95\textwidth]{GABAcomparison2.png}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Independently derived gene sets show similar results 	within a given dictionary.}
  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
   }
  \label{fig:intragree}
\end{figure}

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/EngelmannGO-TEA.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{D. coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D. coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools. 
	\textbf{a)} pantherDB raw results in a tabular format, sorted by p-value. 
	\textbf{b)} TEA raw results in a histogram format, sorted by q-value (lowest on top). 
	Both pantherDB and TEA identify terms associated with neurons (red bars). The two analyses provide complementary, not redundant, information.}
	\label{fig:Dcon}
\end{figure}

% \begin{figure}[h!]
%     \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
% 	\captionsetup{width= 0.95\textwidth}
% 	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
% 	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
% 	  genes normally associated with neuronal precursor tissues as well.
% 	  }
% 	\label{fig:Hsp}
% \end{figure}

>>>>

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes. As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}. These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}). A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. %One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. 
Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.}
	  \label{fig:qvals}
\end{figure}

@@

%gaba
\begin{figure}
  \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
   }
  \label{fig:intragree}
\end{figure}

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools. 
	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
	\textbf{b)} TEA results, sorted by q-value (lowest on top). 
	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.}
	\label{fig:Dcon}
\end{figure}

% \begin{figure}[h!]
%     \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
% 	\captionsetup{width= 0.95\textwidth}
% 	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
% 	  in the posterior lateral ganglion, similarly to \emph{D.~coniospora}. This particular fungus seems to affect
% 	  genes normally associated with neuronal precursor tissues as well.
% 	  }
% 	\label{fig:Hsp}
% \end{figure}
